---
title: "Bio 302 Exam 2020"
author: "Richard J. Telford"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 1) # set default number of decimal places to print in text
```

## Questions

1) Discuss the advantages and challenges of pre-registering an experiment. 

2) A statistical test has a p-value of 0.04.
 - How should this p-value be interpreted? Is it good evidence against the null hypothesis?

3) Explain what autocorrelation is, how it can be detected and how its effect on regression can be controlled for.

4) Model the relationship between total plant biomass in alpine grasslands and summer air temperature in China. Available data are biomass per species per plot. There are ~15 plots in each of four sites. Each site is at a different elevation and has a climate logger.  

Climate data can be downloaded from https://osf.io/34qnr/

Biomass data can be downloaded from https://osf.io/6sfqw/ (you already have these on Rstudio.cloud).

Calculate mean summer air temperatures each site. Use the logger "gradient". The OTC logger is part of another experiment. 

Calculate biomass per plot (you have already done this).

Join the climate data to the biomass data.

Choose and fit a suitable model to find the relation a biomass and mean summer temperature. 

Check the model's assumptions are met.

Report key statistics from your model in a table or in the text.

Make a publication quality plot that shows the relationship between biomass and mean summer temperature.

Write the statistical part of the methods and results sections of a manuscript describing the biomass-climate relationship. You should justify your choice of model.

Write a biological interpretation of your final model.					

Your manuscript should be fully reproducible. That is, I should be able to render your rmarkdown file and generate all results.

## Instructions

Please answer all questions. Please do not google-copy-paste answers - I will check for plagiarism.

Your answers should be fully reproducible and include publication quality graphics where relevant. Submit as html or pdf and include the Rmarkdown file.

If there are problems, please ask me richard.telford@uib.no

# Notes for a solution

# 1) Registered reports

Define pre-registration

Advantages:

 - avoid publication bias (explain what this is, why it is bad etc) 
 - avoid HARKing (define, explain, etc)
 - reduce scope for p-hacking (more explicit pre-registration -> less scope)
 - reviewer feedback can improve experimental design before work done
 - more credible research (because people know it is not result of HARKing etc)
 - forces power tests before analysis - make sure experiment has enough power
 
Disadvantages:

 - takes time before experiment (but much less time afterwards as reviewer 2 cannot ask for an extra experiment)
 - limits flexibility (but can deviate from pre-registered plan but need to justify - lots of flexibility in analysis risks p-hacking)
 - unexpected finding cannot be discussed (they can, just need to be labelled as _post hoc_)
 - unsuitable for some types of work (purely exploratory; urgent work e.g. after an unexpected event that needs researching)

Also:
 
- PARKing possible (pre-registration after results known - this is misconduct)

Useful references:

Chambers, C. 2019. What’s next for Registered Reports? Nature 573: 187–89. https://doi.org/10.1038/d41586-019-02674-6.

# 2) p-value

- Define p-value.
- Significance threshold 
- Explain type-I errors
- False discovery rate
- p = 0.04 is weak evidence against the null. 
- Cannot be properly evaluated without knowing plausibility of alternative hypothesis (implausible alternatives need more compelling evidence).
- effect size is a more useful (but still not sufficient) statistic.

Useful reference:

Colquhoun, D. 2014. An investigation of the false discovery rate and the misinterpretation of p-values. Royal Society Open Science 1:140216. https://doi.org/10.1098/rsos.140216

Also:

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">live video of a statistician trying to keep researchers away from p-values <a href="https://t.co/qd5kHIUEfP">pic.twitter.com/qd5kHIUEfP</a></p>&mdash; 🔥Kareem Carr🔥 (@kareem_carr) <a href="https://twitter.com/kareem_carr/status/1279789002099437569?ref_src=twsrc%5Etfw">July 5, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script> 


# 3) Autocorrelation

- Positive autocorrelation - Observations that are geographically or temporally close are more similar than expected (also phylogenetic autocorrelation).
- Positive autocorrelation causes statistical tests to be too liberal - p-values are too low, confidence intervals are too narrow. Effective number of observations less (perhaps much less) than the number of observations.

Detect:

- graphically with ACF/PACF plots or variograms.
- with Durbin-Watson test (equally spaced observations only)
- test if adding autocorrelation to model (gls or lme) improves the fit (AIC etc)

Control: by adding autocorrelation structure to model (gls or lme) 

Useful reference:

Legendre, P. (1993), Spatial Autocorrelation: Trouble or New Paradigm?. Ecology, 74: 1659-1673. doi:10.2307/1939924

# 4) Biomass analysis

```{r packages, results = "hide", message=FALSE}
library("tidyverse")
library("lubridate")
library("readxl")
library("nlme")
library("broom")
library("here")
```

Multiple ways to process data. ~All code in tidyverse solution below was covered in the course. Non-tidyverse solutions were accepted provided they were clear and reproducible. 

```{r temperature-data}

#Use relative paths. Absolute paths hinder reproducibility
climate_file <- "data/China_2013_2016_AirTemp_month.csv"
#or use here::here
climate_file <- here("data", "China_2013_2016_AirTemp_month.csv")
# here::here is particularly useful in Rmd files which might not be in the project root directory

# import csv file with read.csv or read_csv - latter returns tibble 
clim0 <- read_csv(climate_file)

clim1 <- clim0 %>% 
  #rename month column
  rename(date = month) %>% 
  #filter to keep only logger data
  filter(logger == "gradient") %>%
  #mutating site so 1) has sites in logical order (not alphabetical) and 2) has useful labels
  mutate(site = factor(site, levels = c("H", "A", "M", "L"), labels = c("High alpine", "Alpine", "Mid alpine", "Low alpine")))


#ALWAYS a good idea to plot data - check for dubious values etc
ggplot(clim1, aes(x = date, y = value, colour = site)) + 
  geom_line() + 
  geom_point()
#Note some missing data. Need to decide how to treat this.
#In this case probably OK to just take mean ignoring missing data as this should not give large bias. Would be a problem if e.g. one site only had September data

#calculate summer means
clim <- clim1 %>% 
  select(-variable) %>% 
  #many solutions to filtering out summer data. Simplest is to use lubridate::month. Solutions using regular expressions or tidyr::separate are less transparent and are fragile if date format changes
  # here defining summer as June-September - 4 warmest months. Other definitions valid and probably all highly correlated with each other
  filter(between(month(date), 6, 9)) %>% 
  group_by(site) %>% 
  #Always use TRUE/FALSE not T/F
  summarise(temp = mean(value, na.rm = TRUE))
  
  
clim
```

```{r biomass-data, warning=FALSE}
biomass_file <- here("data", "biomass2015.xls")
#purrr::map_df is short way to import data from multiple sheets/files. lapply %>%  bind_rows also works. do.call("rbind", ...) is a confusing construction best avoided
biomass0 <- map_df(.x = excel_sheets(biomass_file),
                   .f = ~read_excel(biomass_file, sheet = .x))
#this generates a lot of warnings for sheet 4 - not obvious why, data seem fine

biomass <- biomass0 %>% 
  select(site, plot, biomass = production) %>% 
  group_by(site, plot) %>% 
  mutate(site = factor(site, levels = c("H", "A", "M", "L"), labels = c("High alpine", "Alpine", "Mid alpine", "Low alpine"))) %>% 
  summarise(biomass = sum(biomass, na.rm = TRUE)) %>% 
# join biomass to climate data
# merge() also works but is much slower on large datasets and less informative
  left_join(clim, by = "site")

```

```{r quick-biomass-plot}
biomass %>% 
  ggplot(aes(x = temp, y = biomass, fill = site)) +
  geom_boxplot()
```

Grouped data - each site is a cluster. Need to use mixed effect model (or as number of levels in random effect is low, perhaps a ordinary least squares model with site as fixed effect).
Data are continuous, so need model that has a continuous response. Gaussian is the obvious place to start. NB you cannot tell this by looking at a histogram of biomass. You need to look at the residuals. 

```{r model}
model1 <- lme(biomass ~ temp, data = biomass, random = ~1|site)
summary(model1)
```

## Diagnostics

No information on spatial arrangement of plots - cannot check for autocorrelation. 
Testing for AR1 only makes sense if plots are equally spaced along a transect.

### residuals

```{r diagnostic-plot}
plot(model1)
```

Residual variance different at each site - response is heteroskedastic. Possible solutions

- ignore
- allow variance to vary by site
- allow variance to vary by temperature
- change distribution. Perhaps log-normal or gamma with `lme4::glmer` (not shown here). Both these would fail with a plot with zero biomass. Log-normal suggests that variance is related to the amount of biomass.


```{r heteroskedastic}
biomass %>% 
  group_by(site) %>% 
  summarise(variance = var(biomass))
```

```{r}
#update model1 to to let variance vary by site
model2 <- update(model1, weights = varIdent(form = ~1|site))
#update model1 to to let variance vary by temperature
model3 <- update(model1, weights = varFixed(value = ~ temp))
#update model1 to use log-normal
model4 <- update(model1, fixed = log(biomass) ~ temp)
#compare models
AIC(model1, model2, model3)#AIC from model 4 is not comparable
```
I'm using `update` rather than re-writing the whole model. The advantage of this is that if I make any change to the original model (change model type, subset data, change response variable etc), it will automatically be reflected in any new models made with `update` (unless the `update` changes the same aspect of the model).

QQplots

```{r qqnorm}
qqnorm(model2, abline = c(0, 1)) # not great, but not too bad
qqnorm(model2, ~ranef(.)) # too few clusters to be very useful
```


Model 2 is best model with untransformed response and is easier to interpret than log-normal model. Suggests some sites have low variance, some high, but no pattern. With more sites, would be clearer if there is a pattern (e.g. more variance with high biomass)

Is there evidence for a quadratic relationship? We know that at a large scale there will be a quadratic relationship - there will be little biomass anywhere where the mean summer temperature is 50°C! But the warmest site in the data is only `r max(clim$temp)`°C - even colder than Bergen. From an ecological perspective, it seems unlikely that grasslands would have maximum biomass at so low a temperature. From a statistical perspective, the apparent curvature is driven by only one site. With the available data, it is probably prudent to fit a straight line and let the random effect take up the variability. 


## A publication quality plot

There are various ways to plot these data. But any plot showing the relationship between biomass and temperature needs to show the modelled relationship (ideally with confidence intervals - but this is hard with `lme` - https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html).

Here I'm showing two alternatives - I think I prefer the geom_point - gives more emphasis to the trendline rather than the site data.

```{r main-plot}
#predictions
pred <- tibble(
  temp = seq(min(biomass$temp), max(biomass$temp), length.out = 10), 
  biomass = predict(model2, newdata = tibble(temp = temp), level = 0))

ggplot(biomass, aes(x = temp, y = biomass)) + 
  geom_boxplot(aes(fill = site), alpha = 0.5) + #best to use fill rather than colour
  geom_line(data = pred) + 
  #change fill to have meaningful colours
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  #axis labels with units
  labs(x = "Summer temperature °C", y = "Biomass g/0.25 m²") +
  theme_bw()

ggplot(biomass, aes(x = temp, y = biomass)) + 
  geom_point(aes(colour = site), alpha = 0.4, position = position_jitter(height = 0, width = 0.1)) + #jitter *slightly*
  stat_summary(fun.data = "mean_se", geom = "point", mapping = aes(colour = site), size = 3) + #site means
  geom_line(data = pred) + 
  #change fill to have meaningful colours
  scale_colour_brewer(palette = "RdYlBu", direction = -1) +
  #axis labels with units
  labs(x = "Summer temperature °C", y = "Biomass g/0.25 m²") +
  theme_bw()
```


```{r}
model2_coef <- tidy(model2, effects = "fixed")
model2_coef %>% knitr::kable(digits = 2, caption = "Model coefficients")
```

## Methods

Describe what you had done. Cite important packages.

## Results

Make results section reproducible by having inline code to show numbers.

Effect of temperature on biomass is positive (`r model2_coef %>% filter(term == "temp") %>% pull(estimate)` g/0.25 m²/°C), but not significant (p = `r model2_coef %>% filter(term == "temp") %>% pull(p.value)`). Lack of significance is probably because there are so few sites giving few degrees of freedom for the test. A better design would have more sites (perhaps with fewer replicates at each site)

